\myTitle

### 2014-10-29 tex/main-body-text.tex

\section*{Comparison of {\sc structure} and {\sc flock} results}

We used a published dataset of 2596 steelhead, \textit{Oncorhynchus mykiss},
genotyped at 15 microsatellite loci from \citet{Garzaetal_norcal} to compare the programs
{\sc flock} and {\sc structure}.  {\sc flock} 
was run six times using the default setting for all 
run parameters (initial partition = random, number of iterations = 20, number of runs = 50, 
LLOD threshold = 0) for values of $K$ from 2 to 6. The stopping point of 
$K=6$ was chosen after four consecutive values of $K$
yielded no plateau (see below). For each value of $K$ from 2 to 6, the no-admixture, 
uncorrelated-allele-frequency model implemented in {\sc structure} was run 6 times with 
a 50,000 sweep burn-in followed by a 150,000 sweep sample from the posterior. 
All runs were performed on a JNCS D685 with an i7 3.40GHz processor and 8GB
of RAM running 64-bit Windows 7.

@@

\citet{Duc&Tur2012} describe a method for using {\sc flock} to estimate, $K$, the number of
subpopulations. The user must specify both the number of runs (different initial allocations of individuals 
to each cluster) as well as the number of iterations. (In each iteration every individual has the opportunity to be reallocated 
to a different cluster.) At the conclusion of each run, the log-likelihood value for membership to all $K$ clusters is computed
and the log likelihood difference (LLOD), the difference between the cluster with the highest likelihood 
and second highest likelihood, is calculated for each individual. {\sc flock} identifies runs that 
have converged to the same solution by detecting that they have the same mean LLOD score, and, in 
{\sc flock} parlance, a series of runs that have converged to the same solution are termed a ``plateau.'' 
The plateau record is a list of how many runs converged to the same allocation
of individuals. 
\citet{Duc&Tur2012} developed two stopping rules that they use to estimate $K$. In the absence of 
strong prior beliefs about the value of $K$, \citet{Duc&Tur2012} advocate running  {\sc flock} until either a single plateau of
$\geq 6$ runs is reached or four consecutive $K$ values yield no plateaus. If the first stopping condition is satisfied,
the $K$ for which there is a single plateau of $\geq 6$ runs is the point estimate for $K$. Alternatively, 
if multiple plateaus are observed with at least one plateau of $\geq 6$ runs, the largest value of $K$ that satisfies 
this conditon should be taken as the lower bound estimate of $K$. If no plateaus of $\geq 6$ runs are observed, $K$
is declared {\em undecided}.

@@

\subsection*{Results} 
Run times for both programs were relatively short, albeit not trivial. All runs of {\sc flock} 
completed in 1133 minutes. Each of the six groups of runs averaged 188.8 minutes.
All runs of {\sc structure} were completed in 
486 minutes and averaged 80.8 minutes to complete each of the six groups of runs.
The processing time of each algorithm increased as the number or 
putative clusters increases. Increasing
$K$ from of 2 to 6 increased the run time for {\sc flock} by a factor of 2.4, while the run time 
for {\sc structure} increased by a factor of 1.6.

{\sc flock} and {\sc structure} provide similar results in 
terms of how individuals are allocated among runs at low values of $K$, 
while at larger values, {\sc flock} often found, in addition to the solution identified by {\sc structure}, 
multiple other solutions. For all six groups 
of runs with $K = 2$ to 4, {\sc flock} obtained a nearly identical clustering of individuals 
(See supplementary materials) none of which differed substantially from {\sc structure}'s results
(Fig.~\ref{fig:qplots}). Of the `best runs' for the six groups for $K$ = 2, two  
runs yielded identical clustering (same mean LLOD score) while all other runs 
differed by at most the allocation of 13 individuals. As 
$K$ increased so did the number of solutions
that {\sc flock} found (Fig 1). For all groups of runs for $K > 2$, no plateaus were observed---all initial 
allocations resulted in solutions that differed. 
{\sc flock} still produced estimates of the LLOD by individual for the 
'best run', however, it is unclear in these cases what constitutes the 'best run' as all plateau lengths are 
equal to one. Many of these solutions are, however, surprisingly similar to
the partitions identified by {\sc structure}. Both programs idenfity clusters 
which can be attributed to the geographical boundaries identified by \citet{Garzaetal_norcal},
although less reliably so in the case of {\sc flock}.

Despite the similar allocation of individuals into $K$ clusters, the 
plateau analysis to estimate $K$ in {\sc flock} leads to different conclusions among groups of runs, as well as 
one that differs from {\sc structure}. While no plateaus were observed in any of the six groups 
for a $K > 2$, the plateau record varied greatly among each
group for $K = 2$ (plateau sequences: Run1 = 3,2,3,7,4; Run2 = 2,5,2,3,2,3; Run3 = 2,2,2,2,2,5; 
Run4 = 2,3,4,2,4,3; Run5 = 2,2,5,5,3; and; Run6 = 2,2,2,2,2,3,3). The plateau analysis 
procedure suggested by \citet{Duc&Tur2012} leads to
either an estimate of K $\geq$ 2 (Run1) or {\em undecided} as a consequence of no plateau 
lengths being greater than 6. The ($\ln P(D)$) from {\sc structure}
was largest at $K$ = 6 which was further supported by the $q_i$ plots, however,
the $\Delta K$ method supported a $K$ of 3.

\begin{figure*}
\begin{center}
    \includegraphics[width=\textwidth]{images/Figures-Pat/Flock-Fig1.pdf}   % Use this if we want it grayscale -> Flock-Fig1-GreyScale.pdf  
    \caption{Plots of $q_i$ values from {\sc structure} and the normalized likelihood values 
from {\sc flock} made with the program {\sc distruct}. Each horizontal plot represents an individual run of the program {\sc structure}, or the `best run' from a series of runs in {\sc flock}. Within each plot 
$q_i$ values from {\sc structure} and the normalized likelihood values 
from {\sc flock}  are represented by a 
%gray-scale
vertical bar. The numbers in parentheses on the right indicate how many {\sc structure} runs (out of 6, as indicated) reached the same visually indistinguishable solution, or how many groups of {\sc flock} runs (out of 6) had the same visually indistinguishable `best run.'}
    \label{fig:qplots}
\end{center}
\end{figure*}

%\begin{figure}
%\begin{center}
% \includegraphics[width=\columnwidth]{images/Figures-Pat/Fig2.pdf}
%    \caption{Individual \textit{$q_i$} values plotted for each cluster for $K=5$. 
%{\sc flock} estimates of likelihood of assignment to each cluster have less uncertainty
% attributed to them.}
%    \label{fig:qscatter}
%\end{center}
%\end{figure}

To investigate the ability of the arrangement of individuals within the reference 
populations to move around the partition space we varied the number of iterations that 
each run goes through to find the ``optimal'' allocation of individuals. Unfortuantely 
only the reallocation matrix for the `best run' is given as output. Increasing the 
number of iterations did not affect the overall inference of $K$. We 
observed that often during the last iterations individuals would alternate between 
reference groups. Increasing the log likelihood 
difference threshold for reallocation to 0.18 (requiring at least a 1.5 times 
higher likelihood to reallocate the individual) minimized this, but did not affect the 
inference of $K$ (lower bound estimate of $K = 2$).

\section*{Conclusions}
In the past decade, the use of genetic markers to identify population structure and to attempt inference of
the number of genetic clusters ($K$) in a collection has dramatically increased in the fields of ecology, evolution, 
epidemiology, and conservation
genetics. The most commonly used program for doing this is {\sc structure} \citep{Pritchardetal2000,Falushetal2003}, 
a full-featured program with many options that is widely-used, well-tested, and universally regarded as a standard tool amongst molecular ecologists. 
We've shown here that the model incorporated by {\sc flock} is essentially equivalent to the no-admixture model with non-correlated allele frequencies within the program {\sc structure}. In comparing results from {\sc flock} and {\sc structure} 
on a real data set from {\em O. mykiss} we find that the two programs give similar results, only, at values of
$K>4$, {\sc flock}'s results appear to be more variable.

>>>>

\section*{Comparison of {\sc structure} and {\sc flock} results}
Data sets with known $K$ were simulated
(code provided on GitHub repository https://github.com/eriqande/flock-comment)
and evaluated in the programs {\sc flock} and {\sc structure} 
for the known value.The first dataset simulated was composed of five populations of 
sizes 500, 500, 800, 1100, and 1100 comprised of individuals 
with 15 loci and an average pairwise $F_{ST} = 0.03$ between populations. 
The second and third dataset were simulated with population sizes
250, 250, 400, 550, and 550 and average pairwise $F_{ST} = 0.02 and 0.03$
respectively. {\sc flock} was run nine times using the default setting for all 
run parameters (initial partition = random, number of iterations = 20, number of runs = 50, 
LLOD threshold = 0) for the known value of $K$. The no-admixture, 
uncorrelated-allele-frequency model implemented in {\sc structure} was run nine times with 
a 5,000 sweep burn-in followed by a 20,000 sweep sample from the posterior. 
% Looking back some of the runs were under different conditions, not huge differences. 
% But 1000 burn-in and 5,000 sweep look about the same
All runs were performed on a JNCS D685 with an i7 3.40GHz processor and 8GB
of RAM running 64-bit Windows 7.

@@

\citet{Duc&Tur2012} describe a method for using {\sc flock} to estimate, $K$, the number of
subpopulations. The user must specify both the number of runs (different initial allocations of individuals 
to each cluster) as well as the number of iterations. (In each iteration every individual has the opportunity to be reallocated 
to a different cluster.) At the conclusion of each run, the log-likelihood value for membership to all $K$ clusters is computed
and the log likelihood difference (LLOD), the difference between the cluster with the highest likelihood 
and second highest likelihood, is calculated for each individual. {\sc flock} identifies runs that 
have converged to the same solution by detecting that they have the same mean LLOD score, and, in 
{\sc flock} parlance, a series of runs that have converged to the same solution are termed a ``plateau.'' 
The plateau record is a list of how many runs converged to the same allocation
of individuals. 
\citet{Duc&Tur2012} developed two stopping rules that they use to estimate $K$.\citet{Duc&Tur2012} advocate running  {\sc flock} until either a single plateau of
$\geq 6$ runs is reached or four consecutive $K$ values yield no plateaus. If the first stopping condition is satisfied,
the $K$ for which there is a single plateau of $\geq 6$ runs is the point estimate for $K$. Alternatively, 
if multiple plateaus are observed with at least one plateau of $\geq 6$ runs, the largest value of $K$ that satisfies 
this conditon should be taken as the lower bound estimate of $K$. If no plateaus of $\geq 6$ runs are observed, $K$
is declared {\em undecided}.

@@

\subsection*{Results} 
Run times for both programs were relatively short, albeit in the case of {\sc flock} 
non-trivial. {\sc flock} averaged 67.3 minutes
for each run of the first simulated dataset, while each group of 
9 iterations for {\sc structure} averaged 18 minutes. CALCULATE
TIME FOR OTHER SIMULATIONS

{\sc flock} and {\sc structure} provide nearly identical results in 
terms of how individuals are allocated among runs for 
all simulations (Fig.~\ref{fig:Sims}). As $F_{ST}$ decreases 
between populations the solutions that {\sc flock} are more variable. 
The average Euclidean distance between the known $q_i$ value and 
the infered value from each program increases as the genetic 
differentiation among populations decreases. {\sc flock} had an 
smaller mean Euclidean distance over all runs for simulations 
1 and 2, but had much larger variation in simulation 3 (Fig.~\ref{fig:ED}).
%calculate Euclidean distance between flock and structure values.

There was very little variation in the 'best-run' returned by {\sc flock}. 
From the nine different sets of runs for the first simulated dataset, 
three groups of runs ended with 
identical clusters (runs 2,8 \& 9, runs 1 \& 3, and runs 4 \& 7). 
Of the other 31 pairwise comparisions that were not identical
15 differed by the placement of 1 indvidual, 11 differed by the placement
of 2 individuals, and 5 differed by the placement of 3 individuals.

Despite the similar allocation of individuals into $K$ clusters, the 
plateau analysis to estimate $K$ in {\sc flock} did not support $K = 5$. 
For the first simulated dataset plateaus of varioius lengths were observed, 
but none greater than 5 while no plateaus were observed for either second or 
third simulated datasets.

\begin{figure*}
\begin{center}
    \includegraphics[width=\textwidth]{images/Figures-Pat/Simulations_Fig1.pdf}   % Use this if we want it color (I'll make greyscale later)  
    \caption{Plots of $q_i$ values from {\sc structure} and the normalized likelihood values 
of {\sc flock} made with the program {\sc distruct}. Each horizontal plot represents one of nine individual runs of the program {\sc structure}, or the `best run' from a series of nine runs in {\sc flock} at $K=5$. Within each plot 
$q_i$ values from {\sc structure} and the normalized likelihood values 
from {\sc flock}  are represented by a 
%gray-scale
vertical bar. Simulation 1 was composed of populations of 
size 500, 500, 800, 1100, and 1100 with an average pairwise $F_{ST} = 0.03$ between populations. 
The second and third dataset were simulated with population sizes
250, 250, 400, 550, and 550 and average pairwise $F_{ST} =$ 0.02 and 0.03
respectively.}
    \label{fig:Sims}
\end{center}
\end{figure*}

\begin{figure*}
\begin{center}
    \includegraphics[width=0.5\textwidth]{images/Figures-Pat/EucliDist.pdf}   % Use this if we want it color (I'll make greyscale later)  
    \caption{Eucledian distance between true $q_i$ value and inferred value from {\sc structure} and {\sc flock} .}
    \label{fig:ED}
\end{center}
\end{figure*}

\section*{Conclusions}
In the past decade, the use of genetic markers to identify population structure and to attempt inference of
the number of genetic clusters ($K$) in a collection has dramatically increased in the fields of ecology, evolution, 
epidemiology, and conservation
genetics. The most commonly used program for doing this is {\sc structure} \citep{Pritchardetal2000,Falushetal2003}, 
a full-featured program with many options that is widely-used, well-tested, and universally regarded as a standard tool amongst molecular ecologists. 
We've shown here that the model incorporated by {\sc flock} is essentially equivalent to the no-admixture model with non-correlated allele frequencies within the program {\sc structure}. In comparing results from {\sc flock} and {\sc structure} 
on simulated datasets with varying levels of differnetiation among populations
we find that the two programs give similar results.
